Procurement Dynamics: Analyzing Federal Contract Awards to Microsoft

### 2024-12-04 paper/paper.qmd
Final check

The structure of the paper is as follows: [@sec-data] outlines the data sources and variables considered, followed by the model setup and justifications in [@sec-mod]. The results in [@sec-result] presents the key findings of the analysis, with a discussion on the implications. [@sec-discussion] then discusses potential limitations of data, model and suggestions for future research. The appendix provides additional detailed information about models in [@sec-model-details] and methodology in [@sec-meth-details].

@@

```{r}
#| label: fig-amount
#| fig-cap: The histogram shows the distribution of the total amount ($million) of each year's federal contract (since 2020).
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Plot the amount in each year
ggplot(yearly_amount, aes(x = as.factor(Year), y = TotalAmount)) +
  geom_bar(stat = "identity", fill = "darkorange", color = "black", alpha = 0.7) +
  labs(
    title = "Total Contract Amounts by Year for Microsoft",
    x = "Year",
    y = "Total Amount ($million)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8), 
    axis.title.y = element_text(size = 10),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 8),
    plot.title = element_text(size = 12)
    )

@@

```{r}
#| label: tbl-contract
#| tbl-cap: The table summarizes the contract type which has over 25 contracts with the total dollar amount, and average dollar amount per contract. 
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Summarize the data and filter buyers with counts >= 25
summary_table <- cleaned_data %>%
  group_by(ContractType) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 25) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

**Buyer** is a character-type variable representing the federal government department or agency that initiates and awards the contract. [@tbl-buyer] shows the buyer who has over 25 contracts with Microsoft, which identifies key partners and understands the distribution of contracts across sectors. The table highlights that Employment and Social Development Canada (ESDC) and National Defence account for over 50% of Microsoft's total awards.

```{r}
#| label: tbl-buyer
#| tbl-cap: The table summarizes top 10 buyers with contract count, total dollar amount, and average dollar amount per contract. 
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Summarize the data and filter buyers with counts >= 25
summary_table <- cleaned_data %>%
  group_by(Buyer) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 25) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

```{r}
#| label: fig-preparatory
#| fig-cap: The histogram shows the distribution of preparatory period.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

The Full Bayesian Model has an $R^2$ of 0.289, indicating that approximately 28.9% of the variance in the contract amount is explained by the model. In comparison, the Reduced Bayesian Model explains only 8.2% of the variance in the contract amount, suggesting that the inclusion of additional predictors (specific buyers and contract names) improves the model’s explanatory power. However, both models have very low adjusted $R^2$, which suggests that the model is not capturing much of the unexplained variance in the data.

The Full Bayesian Model has a Log-Likelihood of $-15,351.795$ and an ELPD of $-15,519.3$, while the Reduced Model has a Log-Likelihood of $-15,398.022$ and an ELPD of $-15,444.8$. These values suggest that the Full Bayesian Model fits the data slightly better than the Reduced Model, as it produces higher Log-Likelihood and ELPD values.

The Full Bayesian Model has a lower LOOIC ($31,038.7$) and WAIC ($30,992.3$) compared to the Reduced Model, indicating that the Full Bayesian Model is better at out-of-sample predictive performance. Lower values of LOOIC and WAIC are indicative of better model fit and predictive accuracy.

The Full Bayesian Model has a lower RMSE ($2,714,467.04$) compared to the Reduced Model ($2,971,135.35$), indicating better predictive performance. The lower RMSE value suggests that the Full Bayesian Model is more accurate in predicting the total contract amount.

More results about the Bayesian model can be found in @sec-model-details.

@@

## Graphs {#sec-data-visualization}
```{r}
#| label: fig-contract
#| fig-cap: The histogram shows the distribution of performance period, which demonstrates the days between start and end date.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-cor
#| fig-cap: Correlation matrix showing the relationships between numerical predictor variables and the outcome Amount variable
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-box
#| fig-cap: The Boxplots shows the relationship of ContractDays and PhaseDays with ContractType. 
#| fig-subcap: ["Contract Duration in each Contract Type","Phase Duration in each Contract Type"]
#| echo: false
#| warning: false
#| layout-nrow: 2

@@

```{r}
#| label: fig-date
#| fig-cap: Heatmap of Contracts by Four-Month Periods (2020–2024). The heatmap shows the frequency of federal procurement contracts awarded to Microsoft in each four-month period, with darker colors representing higher contract frequencies.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-mobilization
#| fig-cap: The scatter plot shows the relationship between Award Date and Start Date for the top 10 federal procurement contracts types. Each point represents a contract, with colors indicating different contract types. The plot helps to identify trends in the timing between contract award and commencement.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-performance
#| fig-cap: The scatter plot shows the relationship between Start Date and End Date for the top 10 Microsoft’s federal procurement contracts types. Each point represents a contract, and the color reflects the contract type. This visualization provides insights into the duration of the contracts and any deviations from expected timelines.
#| echo: false
#| eval: true
#| warning: false
#| message: false

>>>>

The structure of the paper is as follows: [@sec-data] outlines the data sources and variables considered, followed by the model setup and justifications in [@sec-mod]. The results in [@sec-result] presents the key findings of the analysis, with a discussion on the implications. [@sec-discussion] then discusses potential limitations of data, model and suggestions for future research. The appendix provides additional detailed information about data visualization in [@sec-data-details] models in [@sec-model-details] and methodology in [@sec-meth-details].

@@

@@

```{r}
#| label: tbl-contract
#| tbl-cap: Summarizes the contract type which has over 30 contracts with the total dollar amount, and average dollar amount per contract. 
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Summarize the data and filter buyers with counts >= 30
summary_table <- cleaned_data %>%
  group_by(ContractType) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 30) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

**Buyer** is a character-type variable representing the federal government department or agency that initiates and awards the contract. [@tbl-buyer] shows the buyer who has over 30 contracts with Microsoft, which identifies key partners and understands the distribution of contracts across sectors. The table highlights that Employment and Social Development Canada (ESDC) and National Defence account for over 50% of Microsoft's total awards.

```{r}
#| label: tbl-buyer
#| tbl-cap: Summarizes the buyers who has over 30 contract counts, with total dollar amount and average dollar amount per contract. 
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Summarize the data and filter buyers with counts >= 30
summary_table <- cleaned_data %>%
  group_by(Buyer) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 30) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

@@

The Full Bayesian Model explains 28.9% of the variance in contract amounts, as indicated by the $R^2$ value of 0.289. In comparison, the Reduced Model explains only 8.2% of the variance, highlighting the contribution of additional predictors (e.g., specific buyers and contract names) in improving the model's explanatory power. However, both models show low adjusted $R^2$ values, suggesting that the models do not capture a large portion of the unexplained variance in the data. This points to the need for further refinement, potentially by incorporating additional factors that influence procurement decisions.

The Full Bayesian Model has a Log-Likelihood of $-15,351.795$ and an ELPD of $-15,519.3$, while the Reduced Model's Log-Likelihood is $-15,398.022$ and ELPD is $-15,444.8$. The Full Model performs slightly better, as indicated by its higher Log-Likelihood and ELPD values. Both of these metrics suggest that the Full Model fits the data more accurately, as higher Log-Likelihood and ELPD values correspond to better data fit and prediction accuracy.

The Full Bayesian Model has a lower LOOIC ($31,038.7$) and WAIC ($30,992.3$) compared to the Reduced Model, suggesting better out-of-sample predictive performance. LOOIC and WAIC are used to estimate the out-of-sample prediction accuracy, with lower values indicating better model performance. These results further support the conclusion that the Full Model is superior in predicting contract amounts.

The Full Bayesian Model also outperforms the Reduced Model in terms of RMSE, with a value of $2,714,467.04$ compared to $2,971,135.35$ for the Reduced Model. The lower RMSE indicates that the Full Model is more accurate in predicting the total contract amount, making it a better tool for forecasting future contract values.

More results about the Bayesian model can be found in @sec-model-details. This includes further exploration of model convergence, posterior predictive checks, and alternative specifications that could be explored in future research.

@@

## Graphs {#sec-data-visualization}

```{r}
#| label: fig-amount
#| fig-cap: Total contract amounts awarded to Microsoft ($million) by year (since 2020). The data is grouped by award year, highlighting an increasing trends in Microsoft's federal contracts.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Plot the amount in each year
ggplot(yearly_amount, aes(x = as.factor(Year), y = TotalAmount)) +
  geom_bar(stat = "identity", fill = "darkorange", color = "black", alpha = 0.7) +
  labs(
    title = "Total Contract Amounts by Year for Microsoft",
    x = "Year",
    y = "Total Amount ($million)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8), 
    axis.title.y = element_text(size = 10),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 8),
    plot.title = element_text(size = 12)
    )
```

```{r}
#| label: fig-preparatory
#| fig-cap: Distribution of preparatory period. This shows the frequency of contracts grouped by the number of days spent in the preparatory phase, illustrating the variability in preparatory phase durations across the dataset.
#| echo: false
#| eval: true
#| warning: false
#| message: false

```{r}
#| label: fig-contract
#| fig-cap: Distribution of contract performance phase durations. The histogram shows the frequency of contracts grouped by the number of days in the performance phase, which demonstrates the days between start and end date.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-cor
#| fig-cap: Correlation matrix of numeric variables. This shows a weak negative correlations between $PreparatoryPhase$ and $ContractDays$; and weak positive correlations between predictors and $Amount$.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-box
#| fig-cap: The relationship of ContractDays and PhaseDays by ContractType. 
#| fig-subcap: ["Contract Duration in each Contract Type","Phase Duration in each Contract Type"]
#| echo: false
#| warning: false
#| layout-nrow: 2

@@

```{r}
#| label: fig-date
#| fig-cap: The distribution of federal contracts by Four-Month Periods (2020–2024). The heatmap shows the frequency of federal procurement contracts awarded to Microsoft in each four-month period, with darker colors representing higher contract frequencies.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-mobilization
#| fig-cap: The relationship between Award Date and Start Date for the top 10 federal procurement contracts types. Each point represents a contract, with colors indicating different contract types. The plot helps to identify trends in the timing between contract award and commencement.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-performance
#| fig-cap: The relationship between Start Date and End Date for the top 10 Microsoft’s federal procurement contracts types. Each point represents a contract, and the color reflects the contract type. This visualization provides insights into the duration of the contracts and any deviations from expected timelines.
#| echo: false
#| eval: true
#| warning: false
#| message: false
